A Streaming Algorithm for 2-Center with Outliers in High Dimensions

### 2015-04-26 paper.tex
reverting back to general for 1-center
paper reorganized
Passed to Behnam for completing the proofs, etc.

\usepackage{graphicx}
\usepackage{caption,subcaption}

\usepackage[noend]{algpseudocode}

%----------------------- Environments --------------------------
\newtheorem{corollary}{Corollary} 
\newtheorem{problem}{Problem}
\newtheorem{algorithm}{Algorithm}
\newtheorem{definition}{Definition}

%------------------------ Paper Notations -----------------------------
\DeclareMathOperator*{\argmin}{arg\,min}

%-------------------------- Notations ------------------------------------
\newcommand{\cO}{\ensuremath{{\rm O}}}
\newcommand{\st}{\textsuperscript{\textit{st}} }
\newcommand{\nd}{\textsuperscript{\textit{nd}} }
\newcommand{\nth}{\textsuperscript{\textit{th}} }

@@

\newcommand{\set}[1]{\left\{ #1 \right\}}
\newcommand{\seq}[1]{\left< #1 \right>}
\newcommand{\ceil}[1]{\left\lceil{#1}\right\rceil}
\newcommand{\floor}[1]{\left\lfloor{#1}\right\rfloor}
\newcommand{\card}[1]{\left|{#1}\right|}
\newcommand{\setcomp}[1]{\overline{#1}}
\newcommand{\provided}{\,:\,}

@@

\newcommand{\eps}{\varepsilon}
\newcommand{\bslash}{\!\setminus\!}
\newcommand{\bigOmega}{{\rm\Omega}}
\newcommand{\etal}{{\em et~al.\/}}
\newcommand{\REM}[1]{}
\newcommand{\qed}{}
\newcommand {\backspace}{\vspace{-3em}\\}

@@

%----------------------- Title -----------------------------------------
\title{Streaming $2$-Center with Outliers in High Dimension}
\author{Hamid~Zarrabi-Zadeh\thanks{Department of Computer Engineering, 
	Sharif University of Technology.
	{\tt zarrabi@sharif.edu}}
	\and 
}

%------------------------------ Text -------------------------------------
\begin{document}

@@

\begin{abstract}
In this paper, we study the $2$-center streaming problem with $z$ outliers in a general space, which means finding two congruent balls covering all the points except $z$ outliers in a streaming work-space. Before considering this problem, we provide a  $2$-approximation algorithm for $1$-center with $z$ outliers in a general space. Our algorithm for the $2$-center problem uses an update time and working space polynomial in $d$ and sub-linear in $n$ to calculate a $(1.8+\eps)$-approximation solution.
\end{abstract}

@@

\section{Introduction}
Metric $k$-center problem -- covering a set of point with $k$ congruent balls of minimum radius -- is a fundamental problem in computer science and has various applications in data mining and image processing.
One of $k$\nobreakdash-center problem variants is $k$-center with $z$-outlier which finds $k$ congruent balls with minimum radius covering all but $z$ points of a point set. In real world applications, due to commonness of noisy data, it is important to consider outlier points. Because even a small number of outliers ($\cO(k)$) can increase the radius of $k$\nobreakdash-center problem unboundedly.
In this paper we focus on streaming data model. In the streaming data model, input data arrives as a sequence and there is not enough space to keep the whole data set. Algorithms in this model have limitations in space (typically sub-linear in input size) and processing time per item. This model is specially suitable for massive data sets, due to its sub-linear space complexity \cite{aggarwal2007data}.

The $k$-center problem without outliers (i.e., $z=0$) in general (Euclidean) space can not be approximated with a smaller factor than $2$ ($1.822$), unless $\mathbf{P}=\mathbf{NP}$ \cite{bern1996approximation}.
Gonzalez~\cite{gonzalez1985clustering} showed a $2$-approximation algorithm for general case of the problem.
For small $k$ and $d$, there are more efficient algorithms.
For Euclidean $1$-center in fixed dimensions, Chazelle~\etal~\cite{chazelle1996linear} gave a linear algorithm.
For $2$\nobreakdash-center in the plane the best known algorithm, given by Chan \cite{chan1999more}, runs in $\cO(n \log^2 n \log^2 \log n)$ and linear space.

In streaming data model, Guha \cite{guha2009tight}, in parallel with McCutchen~\etal~\cite{mccutchen2008streaming} gave $(2+\eps)$-approximation for $k$-center without outliers with $\cO((\frac{dk}{\eps}) \log(\frac{1}{\eps}))$ space.
For small $k$, better algorithms are known. For fixed $d$, there exists an $\eps$\nobreakdash-coreset for $1$-center with $\cO(\frac{1}{\eps^{\frac{d-1}{2}}})$ space.
For higher dimensions, Agarwal~\etal~\cite{agarwal2010streaming} gave a $(\frac{1+\sqrt{3}}{2})$-approximation for $1$\nobreakdash-center problem with $\cO((\frac{d}{\eps^3})\log(\frac{1}{\eps}))$ space and and update time, and Chan~\etal~\cite{chan2014streaming} reduced its approximation factor to $1.22$, by an improved analysis.
For $k=2$, Ahn~\etal~\cite{ahn2014computing} gave a $(2+\eps)$-approximation algorithm using $\cO(\frac{d}{\eps})$ space and update time, which was improved for Euclidean spaces by Kim~\etal~\cite{kim2014improved} to a $(1.8+\eps)$-approximation algorithm, using the same space and update time complexity.

For the problem of $k$-center with outliers, Charikar~\etal~\cite{charikar2001algorithms} gives the first offline algorithm with approximation factor $3$ ($4$), where the candidate center points of the centers can be enumerated (otherwise).
For small $k$, there are better results known. $1$-center with $z$ outliers can be solved in $\cO(n \log n + z^{3} n^{\eps})$, for any $\eps > 0$, using Matou{\v{s}}ek's framework \cite{matouvsek1995geometric}.
Agarwal \cite{agarwal2008efficient} gives a randomized $\cO(n z^{7} \log^3 z)$-time algorithm for $2$-center problem in $\IR^2$.

For the problem of $k$-center with $z$ outliers in streaming data model, McCutchen~\etal~\cite{mccutchen2008streaming} (using the idea in Charikar~\etal \cite{charikar2001algorithms} as subroutine) give a $(3+\eps)$-approximation ($(4+\eps)$-approximation, depending on the used subroutine) algorithm using $\cO(\frac{zk}{\eps})$ space.
In special case of $k=1$ in fixed dimensions, there is a streaming $\cO(\frac{z}{\eps^{\cO(d)}})$-space algorithm \cite{agarwal2007space, zarrabi2011almost}.
For Euclidean spaces, Zarrabi-Zadeh \cite{zarrabi2009streaming} gave $(1.22 \times \sqrt{2})$\nobreakdash-approximation for $1$-center with $z$ outliers running in $\cO(dz^3 + (\frac{d}{\eps^3})\log(\frac{1}{\eps}))$ space and poly$(d)$ update time .

In this paper our concern is about $2$-center problem in high dimensional Euclidean space.
The formal definition of problem is as follows. Given sequence $\mathcal{S}$ of points in $d$ dimensional Euclidean space (for arbitrarily large $d$), find two congruent balls of minimum radius, covering all but at most $z$ points. In Section~\ref{sec:2apr}, we give a simple $2$-approximation algorithm for $1$-center problem with space complexity $\cO(z^2 + zd)$. In the next section, we show a $(1.8 + \eps)$-approximation streaming algorithm for $2$-center with $z$ outliers in Euclidean space, and its space complexity is polynomial in $z$ and $d$ and sub-linear in $n$. In Section~\ref{subsec:smaller} we run $z + 1$ parallel instances of the algorithm in order to find the first non-outlier point.
%Kiana: man ino neveshtam vali manzuretun in bud raje be parallel budan az in bishtar begim yani? yekam ziadi tekrar nemishe?

% Kiana: I think first we should mention what others have done and then start declaring our problem. That's why I changed the location of the previous paragraph.

@@

\section{Preliminaries}
\label{sec:pre}
Let $B(c,r)$ denote a ball with center $c$ and radius $r$. We denote by $pq$ the straight line segment between these two points and by $\card{pq}$ the length of this segment. 
Define $\delta$ of two balls $B(c,r)$ and $B'(c',r')$, the minimum distance between $B$ and $B'$ which is equal to $\max (0, \card{cc'}-r-r')$.

Let $B_{2,1}^*(c_1^*, r_2^*)$ and $B_{2,2}^*(c_2^*, r_2^*)$ denote the optimal solution to $2$-center problem with $z$ outliers. By $\delta^*$, denote the distance between $B_{2,1}^*$ and $B_{2,2}^*$, that is, $\delta^* = \max (0, \card{c_1^*c_2^*} - 2r^*)$.

Two congruent balls $B_1$ and $B_2$ with radius $r$ are said to be \emph{$C$-separated}, whenever the distance $\delta$ of two balls is at least $Cr$.

%Let $\mathcal{P}$ be a configuration of $n$ point in $d$-dimensional Euclidean space. A point $c_p$, not necessarily in $\mathcal{P}$, is a \emph{centerpoint} of $\mathcal{P}$, if no open half-space that omits $c_p$ contains more than $\frac{dn}{d+1}$ points of $\mathcal{P}$.

Given an $n$-point set $\mathcal{P}$ in $d$-dimensional Euclidean space, a point $c \in \IR^d$ is called a \emph{centerpoint} of $\mathcal{P}$ if any halfspace containing $c$, contains at least $\ceil{\frac{n}{d + 1}}$ points of $\mathcal{P}$. In other words, any halfspace (or convex set) that avoids a centerpoint can contain at most $\floor{\frac{dn}{d + 1}}$ points of $\mathcal{P}$.

@@

\section{A Simple $2$-Approximation Algorithm for $1$-Center with Outliers}
\label{sec:2apr}

\begin{algorithm}
\label{alg:1center}
\leavevmode
\begin{algorithmic}
\Function{$1$-centerWithOutliers}{$\mathcal{P}, z$}
	\State Let $p_i$ denote the elements of $\mathcal{P}$, for $i \in \set{1, \dots, n}$.
	\ForAll{$i \gets 1, \dots, z+1$}
		\State Assume $p_i$ is a non-outlier point in optimal solution.
		\State $B(c_i, r_i) \gets$ \Call{KnownPoint}{$\mathcal{P}, z, p_i$}
	\EndFor
	\State $i^* \gets \argmin_{i=1}^{z+1} r_i$
	\State \Return $B(c_{i^*}, r_{i^*})$
\EndFunction
\Statex
\Function{KnownPoint}{$\mathcal{P}, z, p$}
	\State $B \gets B(p, 0)$
	\State $Q \gets \emptyset$
	\For{$p'$ in $\mathcal{P}$}
		\If{$p' \notin B$}
			\State Insert $p'$ into $Q$.
			\If{$\card{Q} = z + 1$}
				\State $q$ $\gets$ closest point to $p$ in $Q$
				\State Remove $q$ from $Q$.
				\State $B \gets B(p, \card{pq})$
			\EndIf
		\EndIf
	\EndFor
	\State \Return{B}
\EndFunction
\end{algorithmic}
\end{algorithm}
Function \textproc{KnownPoint} in Algorithm~\ref{alg:1center}, takes a point $p$ as an argument which is guaranteed to be a non-outlier point in the optimal solution. To overcome the lack of knowledge about such a point, function \textproc{$1$\nobreakdash-centerWithOutliers} tries each of the first $z+1$ points of the stream as a candidate for a non-outlier point and generates a ball for each, using function \textproc{KnownPoint}. It returns the ball with the minimum radius. Note that, due to space limitations inherent to streaming data model, the body of for-all loop in function \textproc{$1$-centerWithOutliers} runs in parallel.

\begin{theorem}
Algorithm~\ref{alg:1center} is a $2$-approximation algorithm for $1$-center problem with $z$ outliers, in any dimension.
\end{theorem}

\begin{proof}
Let $B^*(c^*, r^*)$ be the optimal solution. Clearly, there exists a point $p$ among the first $z+1$ points of $\mathcal{P}$, which is not an outlier in the optimal solution (Figure~\ref{fig:1center}). Since $p \in B^*$, then $\card{pp'} \le 2r^*$ for any point $p'\in B^*$. There is at least one point $q'$ among the $(z+1)$-furthest points from $p$ that is not an outlier in the optimal solution. Thus $\card{pq} \le \card{pq'} \le 2r^*$ and Algorithm~\ref{alg:1center} returns a valid solution of radius at most $2r^*$. So the proof is complete.
\end{proof}

%
%\begin{figure}[h]
%	\centering
%	\includegraphics[width=17em]{figs/one-center}
%	\caption{Output of Algorithm~\ref{alg:1center}}
%	\label{fig:1center}
%\end{figure}

% Behnam: Aks be nazaram gooya nist ziad. va inke jaii behesh refrence dade nashode!
% Kiana: hagh ba behname. akse bedune refrence nadarim asan. montaha man benazaram kheili ehtiaji nis behesh asan 
% Sahand: hazf konim ya pishnehadi baray e behboodesh darid?
% Kiana: Man be nazaram vaghean chon kheili az idehaye basic ham baraye gheire streaming hamine va osulan idash yekam badihitowre hazf beshe. felan commentesh kardam age nazaretun chize digei bud avazesh konin.

The space complexity of Algorithm~\ref{alg:1center} is $\cO(z^2 + zd)$, the update time is $\cO (z \log z + zd)$ and the query time is $\cO (z)$.
If a non-outlier point $p$ is known then the space, update time and query time complexities of Algorithm~\ref{alg:1center} can be reduced by a factor of $z$.

@@

\section{The $2$-Center Problem with $z$ Outliers}

In all the algorithms given in this section it is assumed that $p_1$ is a non-outlier point. It is not hard to see that, this limitation can be overcome by considering $\cO(z)$ parallel instances of the algorithm, similar to Algorithm~\ref{alg:1center}.
% Behnam: Move this to upper section, this assumption can be used for previous algorithm
%kiana: Here is better. Since we do not have p_1 in section 4.2

\begin{theorem}
For any fixed $d > 1$, our single-pass data stream algorithm whose working space and update time are polynomial in $d$ and sub-linear in $n$, returns two centers that guarantees a $(1.8 + \eps)$-approximation for the Euclidean $2$-center problem with outliers.
\end{theorem}

In Section \ref{subsec:smaller}, we give a description of our algorithm for a given $r' > 0$ for the case where $\delta^* < C r^*$ and $1.2r^* \le r' < (1.2 + 2\eps/3)r^*$, which returns a $(1.8 + \eps)$-approximate solution. We explain how to find such an $r'$ and present a full description of our algorithm for the case where $\delta^* < C r^*$ in Subsection~\ref{subsec:findr}. Then we inspect the case where $\delta^* \ge C r^*$ in Subsection~\ref{subsec:bigger}.

>>>>

\usepackage{caption,subcaption}
\usepackage{booktabs,multirow}

\usepackage{algorithm}
\usepackage[noend]{algorithmic}

\usepackage{cleveref}

%------------------------ Paper Notations -----------------------------

\newcommand{\rc}{r}
\newcommand{\cp}{c_p}
\newcommand{\dz}{(d + 1)(z + 1)}
\newcommand{\Buffer}{\ensuremath{\text{Buffer}}}

\renewcommand{\O}{\ensuremath{{O}}} 
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\nd}{\nobreakdash-}

%----------------------- Page Style -----------------------------------------

\renewcommand{\arraystretch}{1.3}

%----------------------- Environments --------------------------

\newtheorem{problem}{Problem}
\newtheorem{definition}{Definition}

%------------------------ Algorithm -----------------------------

\newcommand{\textproc}{\textsc}
\newcommand{\Call}[2]{\textsc{#1}(#2)}
\renewcommand{\algorithmicforall}{{\bf for each}}

%-------------------------- Notations ------------------------------------

@@

\newcommand{\set}[1]{\left\{ #1 \right\}}
\newcommand{\seq}[1]{\left< #1 \right>}
\newcommand{\ceil}[1]{\left\lceil{#1}\right\rceil}
\newcommand{\floor}[1]{\left\lfloor{#1}\right\rfloor}
\newcommand{\card}[1]{\left|{#1}\right|}
\newcommand{\len}[1]{\|{#1}\|}
\newcommand{\setcomp}[1]{\overline{#1}}
\newcommand{\provided}{\,|\,}

@@

\newcommand{\eps}{\varepsilon}
\newcommand{\bslash}{\!\setminus\!}
\newcommand{\bigOmega}{{\rm\Omega}}
\newcommand{\etal}{{\em et~al.\/}}
\newcommand{\REM}[1]{}
\newcommand{\qed}{}
\newcommand{\backspace}{\vspace{-3em}\\}

@@

%----------------------- Title -----------------------------------------

\title{A Streaming Algorithm for 2-Center with Outliers in High Dimensions}
%\title{Computing 2-Center with Outliers on High-Dimensional Data Streams}
\author{Behnam~Hatami\thanks{Department of Computer Engineering, 
	Sharif University of Technology.
	{\tt hatami@ce.sharif.edu}}
	\and 
	Hamid~Zarrabi-Zadeh\thanks{Department of Computer Engineering, 
	Sharif University of Technology.
	{\tt zarrabi@sharif.edu}}
}

%------------------------------ Text -------------------------------------

\begin{document}

@@

\begin{abstract}
We study the Euclidean 2-center problem with outliers in the data stream model. 
Given a stream of points in arbitrary $d$ dimensions, the goal is to find two congruent balls 
of minimum radius covering all but $z$ points. 
We provide a $(1.8+\eps)$-approximation streaming algorithm for the problem, 
improving upon the previous $(4 + \eps)$-approximation algorithm available for the problem.
The space complexity and update time of our algorithm is $\poly(d, z, {1 \over \eps})$,
independent of the size of the stream.
\end{abstract}

@@

\section{Introduction}
The $k$-center problem---covering a set of points 
using $k$ congruent balls of minimum radius---is a fundamental problem,
arising in many applications 
such as data mining, unsupervised learning, statistics, and image processing.
In real-world applications, where input data is often noisy, 
it is very important to consider outliers, 
as even a small number of outliers can greatly affect the quality of the solution.
The $k$-center problem is particularly very sensitive to outliers,
and even a constant number of outliers can increase the radius of the $k$-center unboundedly.
Therefore, it is natural to consider the following generalization of
the the $k$-center problem: % which we call \emph{$k$-center with $z$ outliers}:
given a set $P$ of $n$ points in arbitrary $d$ dimensions, 
find $k$ congruent balls of minimum radius 
to cover at least $n - z$ points of $P$.
%One of $k$-center problem variants is $k$-center with $z$-outlier which finds $k$ congruent balls with minimum radius covering all but $z$ points of a point set.

In this paper, we focus on the \emph{data stream model}
where only a single pass over the input data is allowed,
and we have only a limited amount of working space available.
%which is typically sublinear in the input size.
This model is specially useful for processing large data sets,
as it does not need the entire data set to be stored in memory.

%In this model, input data arrives as a sequence over time, 
%%only one pass over the input data is allowed, 
%and we have only a limited amount of working space available, 
%which is typically sublinear in the input size.
%and thus, we cannot keep the entire data set in memory.
%This makes designing streaming algorithms very challenging,
%%as we can have only one pass over the input data, 
%%and must decide on the fly which data to keep in our limited working space,
%as we cannot keep the entire data set in memory,
%and must be still able to produce a solution comparable to 
%that of an offline algorithm that has unlimited access to the whole data set.

The Euclidean $k$-center problem has been extensively studied in the literature.
%There is a considerable amount of work on the geometric $k$-center problem.
If $k$ is part of the input,
the problem in known to be NP-hard in two or more dimensions~\cite{fowler1981optimal},
and is even hard to approximate to within a factor close to 2, 
unless P$\,=\,$NP~\cite{feder1988optimal}.
Factor-2 approximation algorithms are available for the problem 
in any dimension~\cite{gonzalez1985clustering,feder1988optimal}.
%The $k$-center problem without outliers (i.e., $z=0$) in general (Euclidean) space can not be approximated with a smaller factor than $2$ ($1.822$), unless $\mathbf{P}=\mathbf{NP}$~\cite{bern1996approximation}.
%Gonzalez~\cite{gonzalez1985clustering} showed a $2$-approximation algorithm for general case of the problem.
For small $k$ and $d$, %there are more efficient algorithms.
better solutions are available.
The 1-center problem in fixed dimensions is known to be LP-type 
and can be solved in $O(n)$ time~\cite{chazelle1996linear}.
For 2-center in the plane, the current best algorithm
runs in $\O(n \log^2 n \log^2 \log n)$ time~\cite{chan1999more}.

For $k$-center with outliers, Charikar~\etal~\cite{charikar2001algorithms} 
gave the first algorithm with an approximation factor of $3$, 
that works in any dimension.
%which is improvable to 3, if all candidate center points can be enumerated.
%depending on whether the candidate center points can be enumerated or not, respectively.
Better results are known for small $k$ in the plane.
The 1-center problem with $z$ outliers in the plane can be solved 
in $\O(n \log n + z^{3} n^{\eps})$ time, for any $\eps > 0$, 
using Matou{\v{s}}ek's framework~\cite{matouvsek1995geometric}.
Agarwal~\cite{agarwal2008efficient} gave a randomized 
$\O(n z^{7} \log^3 z)$-time algorithm for 2-center with $z$ outliers in the plane.

In the streaming model, McCutchen~\etal~\cite{mccutchen2008streaming} 
and Guha~\cite{guha2009tight} presented algorithms to maintain $(2+\eps)$-approximation 
to the $k$-center problem in %in any metric space using 
$\O(\frac{kd}{\eps} \log \frac{1}{\eps})$ space.
For $k=1$, %better algorithms are known. % in high dimensions.
%a $1.5$-approximation in high dimensions was presented by Zarrabi-Zadeh and Chan~\cite{zarrabi2006simple}.
a factor-$((1+\sqrt{3})/{2})$ approximation 
was presented by Agarwal and Sharathkumar~\cite{agarwal2010streaming} 
in high dimensions, using $O(d)$ space. 
The approximation factor was later improved 
%to $(1+\sqrt{3})/{2}$ by Agarwal and Sharathkumar~\cite{agarwal2010streaming} 
to $1.22$ by Chan and Pathak~\cite{chan2014streaming}.
%If the dimension is fixed, a $(1 + \eps)$-approximation
%can be maintained in $\O({k}/{\eps^{((d-1)/2)}})$ space 
%using the notion of core-sets~\cite{zarrabi2008core}., 
%Ahn~\etal~\cite{ahn2014computing} gave a $(2+\eps)$-approximation algorithm using $\O(\frac{d}{\eps})$ space and update time, which was improved for Euclidean spaces by 
For $k=2$, Kim~\etal~\cite{kim2014improved} 
have recently obtained a $(1.8+\eps)$-approximation  
using $\O(\frac{d}{\eps})$ space and update time.

For $k$-center with $z$ outliers in the streaming model, 
McCutchen~\etal~\cite{mccutchen2008streaming} 
%(using the idea in Charikar~\etal~\cite{charikar2001algorithms} as subroutine) 
gave a $(4+\eps)$-approximation 
%(or $(3+\eps)$-approximation, if all center points can be enumerated) 
algorithm using $\O(\frac{zk}{\eps})$ space.
When dimension is fixed, a $(1 + \eps)$-approximation to 1-center with outliers
can be maintained in $\O({z}/{\eps^{((d-1)/2)}})$ space 
using the notion of robust $\eps$-kernels~\cite{agarwal2007space, zarrabi2011almost}.
For 1-center with outliers in high dimensions, 
Zarrabi-Zadeh and Mukhopadhyay~\cite{zarrabi2009streaming} 
gave a $(\sqrt{2}\alpha)$-approximation, 
where $\alpha$ is the approximation factor of the underlying algorithm for maintaining 1-center. 
Combined with the $1.22$-approximation algorithm of Chan and Pathak~\cite{chan2014streaming},
it yields an approximation factor of $(\sqrt{2} \times 1.22) < 1.73$ 
using $O(d^3z)$ space and $\poly(d,z)$ update time.
%In fixed dimensions, there is a streaming $\O(\frac{z}{\eps^{\O(d)}})$-space algorithm~\cite{agarwal2007space, zarrabi2011almost}.

\begin{table}[t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{\begin{tabular}[c]{@{}c@{}} \ \  Problem \ \ \end{tabular}} & \multicolumn{2}{c|}{Approximation Factor} \\ \cline{2-3} 
 & Without Outliers & With Outliers \\ \hline
1-Center & 1.22~\cite{chan2014streaming} & 1.73~\cite{zarrabi2009streaming} \\ \hline
2-Center & $1.8 + \eps$~\cite{kim2014improved} & \textbf{$1.8 + \eps$} [Here]  \\ \hline
$k$-Center & $2 + \eps$~\cite{guha2009tight,mccutchen2008streaming} & $4 + \eps$~\cite{mccutchen2008streaming}  \\ \hline
\end{tabular}
\caption {Summary of the streaming algorithms 
for $k$-center with and without outliers in high dimensions.}
\label{table:summary}
\end{table}

\paragraph{Our result}
In this paper, we study the 2-center problem with outliers in high dimensions.
We present a streaming algorithm for the problem that achieves
an approximation factor of $1.8 + \eps$, for any $\eps > 1$,
%using $O(d^3z/\eps)$ space.
using $\poly(d, z, {1 \over \eps})$ space and update time.
This improves the current streaming algorithm available for the problem
which has an approximation factor of $4+\eps$.
%Our work not only improves the best previous algorithm available for the case of 2-center,
%but is also counterintuitive in the sense that its 
The approximation factor of our algorithm matches
that of the best streaming algorithm for the 2-center problem without outliers.
This is somewhat surprising, considering that 
the best current approximation factors for streaming $k$-center with and without outliers 
differ by a factor of $\sqrt{2}$ for $k=1$,
and differ by a factor of 2  for general~$k$. %the 1-center problem.
See \Cref{table:summary} for a comparison.

To obtain our result, we have used a combination of several ideas
including parallelization construction, far/close ball separation,
and keeping a lower bound on the radius and distance of the optimal balls.
Our algorithm also utilizes the ideas used in~\cite{kim2014improved} 
for the 2-center problem with no outliers.
Nevertheless, our algorithm is much detailed and uses many new ingredients.
Indeed, handling outliers makes the $k$-center problem much harder,
as we not only need to find clusters with minimum radius, 
but we also need to decide which subset of points to cluster.
This is in particular more challenging in the streaming model,
where we only have a single pass over the input, and we must decide on the fly
which point can be an outlier, and which one can be safely dropped as a non-outlier point.

\REM{
The rest of this paper is organized as follows.
In Section~\ref{sec:1-center}, we give a simple $2$-approximation algorithm for 
the 1-center problem with outliers that uses $\O(z^2 + zd)$ space.
In the next section, we show a $(1.8 + \eps)$-approximation streaming algorithm for 2-center with $z$ outliers in high dimensions. 
%In Section~\ref{subsec:smaller} we run $z + 1$ parallel instances of the algorithm in order to find the first non-outlier point.
}

@@

\section{Preliminaries}
\label{sec:pre}
Let $B(c,r)$ denote a ball of radius $r$ centered at $c$. 
Given two points $p$ and $q$, we denote by $pq$ the straight line segment between $p$ and $q$,
and denote by $\len{pq}$ the length of this segment. 
Given two balls $B(c,r)$ and $B'(c',r')$, 
we define 
$\delta(B, B') = \max \set{0, \len{cc'}-r-r'}$
to be the \emph{distance} between $B$ and $B'$.
Two congruent balls $B_1$ and $B_2$ with radius $r$ are said to be \emph{$\alpha$-separated}, 
if $\delta(B_1, B_2) \geq \alpha r$.

Given an $n$-point set $P$ in $d$-dimensions,
a point $c \in \IR^d$ is called a \emph{centerpoint} of $P$,
if any halfspace containing $c$ contains at least $\ceil{{n}/({d + 1})}$ points of $P$. 
%In other words, any halfspace (or convex set) that avoids a centerpoint can contain at most $\floor{\frac{dn}{d + 1}}$ points of $P$. 
It is well-known that any finite set of points in $d$-dimensional space 
has a centerpoint~\cite{danzer1963helly}. 
The following observation is a corollary of this fact.
%~\cite{edelsbrunner1987algorithms}.

\begin{obs}
\label{obs:omitting-centerpoint}
	Given a set $P$ of $k(d+1)$ points in $d$-dimensional space, 
	there is a point $c$, not necessarily in $P$, 
	such that any convex object that does not cover $c$, 
	leaves at least $k$ points of $P$ uncovered.
\end{obs}

Given a point set $P$ and a point $p \in P$,
the \emph{$k$-furthest point from $p$}
is the $k$-th point of $P$ when points are sorted
in their decreasing distance from~$p$.

We assume the standard word-RAM model of computation. 
Each coordinate value takes a unit of space.
Thus, a point in $d$ dimensions takes $\O(d)$ space,
and basic operations on the points take $\O(d)$ time.

@@

\section{A 2-Approximation for 1-Center with Outliers}
\label{sec:1-center}

To warm up, we present a simple $2$-approximation streaming algorithm
for the 1-center problem with outliers.
It utilized a parallelization technique that 
will be used extensively in the rest of the paper.
The pseudocode is provided in \Cref{alg:1-center}.
The algorithm receives as input a stream of points, $P$,
the number of outliers, $z$, 
and a special point $c \in P$ which is guaranteed to be 
a non-outlier in the optimal solution.
It returns a ball $B$ covering all but at most $z$ points of $P$.
%We will show later how to remove this assumption.

%\begin{algorithm}
%\caption{\sc 1-Center$(P, z)$} 
%\algsetup{indent=1.5em}
%\label{alg:1-center}
%\begin{algorithmic}[1]
%
%	%\STATE Let $p_i$ denote the elements of $P$, for $i \in \set{1, \dots, n}$.
%	\STATE let $p_1, \ldots, p_n$ be the points of $P$
%	\FOR{$i = 1, \dots, z+1$}
%		%\STATE move $p_i$ to the front of $P$ %is a non-outlier point in optimal solution.
%		\STATE $B_i \gets$ \Call{1-Center-Special}{$P, z,p_i$}
%	\ENDFOR
%	%\STATE $i^* \gets \argmin_{i=1}^{z+1} r_i$
%	\RETURN smallest ball among $\set{B_1, \ldots, B_{z+1}}$
%
%\end{algorithmic}
%\end{algorithm}

\begin{algorithm}
\caption{\sc 1-Center$(P, z, c)$} 
\algsetup{indent=1.5em}
\label{alg:1-center}
\begin{algorithmic}[1]
	%\STATE {$\triangleright$ $c$ is a non-outlier point in the optimal solution}
	\STATE $B \gets B(c, 0)$
	\STATE $Q \gets \emptyset$
	\FORALL {$p$ in $P$}
		\IF {$p \notin B$}
			\STATE insert $p$ into $Q$
			\IF {$|Q| = z + 1$}
				\STATE $q$ $\gets$ closest point to $c$ in $Q$
				\STATE remove $q$ from $Q$
				\STATE $B \gets B(c, \len{cq})$
			\ENDIF
		\ENDIF
	\ENDFOR
	\RETURN {B}
\end{algorithmic}
\end{algorithm}

%The function \textproc{1-Center-Special} in \Cref{alg:1-center-special}, 
%takes as input a point $c$ which is guaranteed to be a non-outlier point in the optimal solution. 
%To overcome the lack of knowledge about such a point, 
%function \textproc{1-Center} in \Cref{alg:1-center}
%tries each of the first $z+1$ points of the stream as a candidate for a non-outlier point and 
%generates a ball for each, using function \textproc{1-Center-Special}. 
%It returns the ball with the minimum radius. 
%Note that, in the streaming model, %due to space limitations inherent to streaming data model, 
%the body of the for loop in function \textproc{1-Center} runs in parallel.
%To run function \textproc{1-Center} in the data stream model,
%we run the body of the for loop in $z+1$ parallel streaming instances.

\begin{theorem} \label{thm:1-center}
	Algorithm~\ref{alg:1-center} computes a $2$-approximation to
	the 1-center problem with $z$ outliers,
	provided that a non-outlier point in the optimal solution in given.
\end{theorem}

\begin{proof}
Let $B^*(c^*, r^*)$ be the optimal solution,
and $c$ be a non-outlier in the optimal solution.
Since $c$ is covered by $B^*$,
for all points $p \in B^*$, we have
$\len{cp} \le \len{cc^*} + \len{c^*p} \le 2r^*$.
Among the $z+1$ points furthest from $c$,
there is at least one point $q$ which is not outlier in the optimal solution,
and therefore, is contained in $B^*$
(see \Cref{fig:1center}). 
Thus, $\len{cq} \le 2r^*$,
and therefore, the ball $B(c, \len{cq})$ returned by Algorithm~\ref{alg:1-center} 
is a 2-approximation. 
\end{proof}

\begin{figure}[t]
	\centering
	\includegraphics[width=12em]{figs/one-center}
	\caption{Proof of \Cref{thm:1-center}}
	\label{fig:1center}
\end{figure}

\noindent
Algorithm~\ref{alg:1-center} 
assumes that a point $c$ is known in advance
which is guaranteed to be a non-outlier in the optimal solution. 
To overcome the lack of knowledge about such a point, 
we run $z+1$ instances of \Cref{alg:1-center} in parallel,
each of which is given as $c$ one of the first $z + 1$ points in the stream.
Clearly, there exists a point among the first $z+1$ points of $P$
which is not an outlier in the optimal solution.
Therefore, the smallest ball among the $z+1$ balls computed in parallel
is always within factor~2 of the optimal solution. % by \Cref{thm:1-center-special}.
Since the space complexity of Algorithm~\ref{alg:1-center} for one instance is $\O(zd)$, 
and its update time is $\O (zd \log z)$, % and the query time is $\O (z)$.
we get the following result.

\begin{theorem} \label{thm:1-center-stream}
	Given a stream of points in $d$ dimensions,
	we can maintain a 2-approximation to the 1-center with $z$ outliers
	in $\O(z^2d)$ space and $\O (z^2d\log z)$ update time.
\end{theorem}

%function \textproc{1-Center} in \Cref{alg:1-center}
%tries each of the first $z+1$ points of the stream as a candidate for a non-outlier point and 
%generates a ball for each, using function \textproc{1-Center-Special}. 
%It returns the ball with the minimum radius. 
%Note that, in the streaming model, %due to space limitations inherent to streaming data model, 
%the body of the for loop in function \textproc{1-Center} runs in parallel.
%To run function \textproc{1-Center} in the data stream model,
%we run the body of the for loop in $z+1$ parallel streaming instances.
%
%The space complexity of Algorithm~\ref{alg:1center} is $\O(z^2 + zd)$, the update time is $\O (z \log z + zd)$ and the query time is $\O (z)$.
%If a non-outlier point $p$ is known then the space, update time and query time complexities of Algorithm~\ref{alg:1center} can be reduced by a factor of $z$.

@@

\section{The 2-Center Problem with Outliers}
\label{sec:2-center}

In this section, we provide a $(1.8 + \eps)$-approximation algorithm
for the 2-center problem with outliers.
In all algorithms presented in this section,
we assume that the first point of the stream, $p_1$, is non-outlier. 
This assumption can be easily removed by considering $z + 1$ parallel instances of the algorithm, 
similar to what we did in \Cref{sec:1-center}.

Let $B_1^*$ and $B_2^*$ be the optimal balls
in an optimal solution to the 2-center problem with $z$ outliers on a point set $P$.
We denote by $r^*$ the optimal radius,
and by $\delta^*$ the distance between $B_1^*$ and $B_2^*$.
To prove our main result, we distinguish between two cases.
In \Cref{subsec:bigger}, we address the case where $\delta^* > \alpha r^*$, 
for some constant $\alpha$ to be fixed later.
(It will be turn out that $\alpha = 14$ is a proper choice.)
We then present in \Cref{subsec:smaller} 
our algorithm for the case of $\delta^* \leq \alpha r^*$.

%We inspect the case where $\delta^* \ge \alpha r^*$ in Subsection~\ref{subsec:bigger}.
%give a description of our algorithm for a given $r > 0$ for the case where $\delta^* < \alpha r^*$ and $1.2r^* \le r < (1.2 + 2\eps/3)r^*$, which returns a $(1.8 + \eps)$-approximate solution. We explain how to find such an $r$ and present a full description of our algorithm for the case where $\delta^* < \alpha r^*$ in Subsection~\ref{subsec:findr}.

% -------------------------------- First Case -----------------------------------
